---
title: "Effect of an organic food intervention on biomarkers of exposure to lead and cadmium in primary school children: A cluster-randomized cross-over trial - Part 2"
author: "Nikolaos Efthymiou"
date: "18/08/2022"
output:
  html_document: default
editor_options:
  chunk_output_type: console
---

# Regression analysis (mixed effects models)

This script includes the regression analysis (Part 2).
The output of the Part 1 script needs to be created (organiko_metals_part1.rds)

```{r initialization, include=FALSE}

## Prepare workspace & install libraries
rm(list = ls(all = TRUE))

ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) 
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}

# usage
packages <- c("readxl","tidyverse","magrittr","knitr","lubridate",
              "lmerTest","lme4","Matrix","sjPlot","broom.mixed","gridExtra","easystats","merDeriv")

ipak(packages)


### Other libraries
###
library(conflicted)  # An Alternative Conflict Resolution Strategy
conflict_prefer("filter", "dplyr")
conflict_prefer("lmer", "lmerTest")
knitr::opts_chunk$set(echo=FALSE, warning = FALSE, results='asis')
# reproducible results
set.seed(1)
```

```{r linear mixed effect models - prepare database, echo=FALSE}
model_data_log_centered <- readRDS("organiko_metals_part1.rds")
```

## Main analysis

### Primary outcomes

Selection of models for the analysis of main outcomes:

Mixed Effect Model:


 + Outcome: Pb/Cd
 + Variable of interest: organic treatment (vs conventional)
 + Random part: participants
 + Adjustment for: baseline levels of outcome (Pb/Cd), time (days from treatment), creatinine
 + Creatinine adjustment:
    + Pb/Cd divided by creatinine
    + raw measurements of Pb/Cd and creatinine as fixed effect
 + Interaction term: time*phase
    + if p-value>0.1 remove from the model  

```{r set1}
# set1
models <- tibble(Response     = c("adj_Pb1000_TR","Pb1000_TR","adj_Cd1000_TR","Cd1000_TR"),
                 Predictors_1 = c("adj_Pb1000_BL","Pb1000_BL + creatinine","adj_Cd1000_BL","Cd1000_BL + creatinine"),
                 Predictors_2 = "DaysfromTreatment + phase",
                 Predictors_3 = c("phase*DaysfromTreatment"),
                 Predictors_4 = "(1 | ID)")

model_list <- list()


for (i in 1:nrow(models)) {
  # formula with interaction term
  form <- as.formula(models %$% str_c(Response[i], " ~ ", Predictors_1[i], " + ", Predictors_2[i], " + ", Predictors_3[i], " + ", Predictors_4[i]))
  model <- lmer(form, model_data_log_centered)
  
  # if p.value of interaction term >0.1 then remove it from formula
  if (model %>% tidy() %>% filter(str_detect(term,":")) %$% p.value > 0.1) {
    form <- as.formula(models %$% str_c(Response[i], " ~ ", Predictors_1[i], " + ", Predictors_2[i], " + ", Predictors_4[i]))
    model <- lmer(form, model_data_log_centered)    
  }
  
  model_list[[i]] <- model
}

tab_model(model_list , show.icc = TRUE, digits = 3, show.obs = TRUE, show.re.var = TRUE)

# fdr adjustment
model_list %>% 
map(~tidy(.)) %>% bind_rows() %>% filter(!term=="(Intercept)") %>%  filter(!str_detect(term, "sd")) %>% 
  mutate(p.adjust = p.adjust(p.value,method="BH"),.after=p.value) %>% 
  mutate(Response = c("adj_Pb1000_TR","adj_Pb1000_TR", "adj_Pb1000_TR","adj_Pb1000_TR",
  "Pb1000_TR", "Pb1000_TR", "Pb1000_TR", "Pb1000_TR","Pb1000_TR",
"adj_Cd1000_TR", "adj_Cd1000_TR", "adj_Cd1000_TR", "adj_Cd1000_TR", 
"Cd1000_TR", "Cd1000_TR","Cd1000_TR","Cd1000_TR"),.before=term) %>%
  arrange(p.adjust) %>%
  select(!group) %>% 
  kable()

ggplot(model_data_log_centered, aes(DaysfromTreatment ,adj_Pb1000_TR,colour = phase)) +  
  geom_point(na.rm = TRUE) +
  geom_smooth(formula = y ~ x,method = "lm", se = FALSE) +
  labs(y = "Pb (ng/g cr)", x = "Time (days)") + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))
  

ggplot(model_data_log_centered, aes(DaysfromTreatment ,Pb1000_TR,colour = phase)) +  
  geom_point(na.rm = TRUE) +
  geom_smooth(formula = y ~ x,method = "lm", se = FALSE) +
  labs(y = "Pb (ng/L)", x = "Time (days)") + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

ggplot(model_data_log_centered, aes(DaysfromTreatment ,adj_Cd1000_TR,colour = phase)) +  
  geom_point(na.rm = TRUE) +
  geom_smooth(formula = y ~ x,method = "lm", se = FALSE) +
  labs(y = "Cd (ng/g cr)", x = "Time (days)") + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))

ggplot(model_data_log_centered, aes(DaysfromTreatment ,Cd1000_TR,colour = phase)) +  
  geom_point(na.rm = TRUE) +
  geom_smooth(formula = y ~ x,method = "lm", se = FALSE) +
  labs(y = "Cd (ng/L)", x = "Time (days)") + 
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
        panel.background = element_blank(), axis.line = element_line(colour = "black"))
```

### Secondary outcomes

Selection of models for the analysis of secondary outcomes:

Mixed Effect Model:


 + Outcome: 8-iso-PGF2a/MDA/8-OHdG
 + Variable of interest: Cd/Pb
 + Random part: participants
 + Adjustment for: baseline levels of outcome (8-iso-PGF2a/MDA/8-OHdG), time (days from treatment), creatinine, age, sex 
 + Creatinine adjustment: 
    + 8-iso-PGF2a/MDA/8-OHdG divided by creatinine
    + raw measurements of 8-iso-PGF2a/MDA/8-OHdG and creatinine as fixed effect 


#### Creatinine adjusted 8-iso-PGF2a/MDA/8-OHdG    


```{r set2}
# set2
models <- tibble(Response     = c("adj_Ohdg_TR","adj_Mda1000_TR","adj_Isopgf2a1000_TR","adj_Ohdg_TR","adj_Mda1000_TR","adj_Isopgf2a1000_TR"),
                 Predictors_1 = c("adj_Ohdg_BL","adj_Mda1000_BL","adj_Isopgf2a1000_BL","adj_Ohdg_BL","adj_Mda1000_BL","adj_Isopgf2a1000_BL"),
                 Predictors_2 = c("adj_Cd1000_TR","adj_Cd1000_TR","adj_Cd1000_TR","adj_Pb1000_TR","adj_Pb1000_TR","adj_Pb1000_TR"),
                 Predictors_3 = "Sex + age_baseline + DaysfromTreatment",
                 Predictors_4 = "(1 | ID)")

model_list_osi1 <- list()


for (i in 1:nrow(models)) {
  # formula with interaction term
  form <- as.formula(models %$% str_c(Response[i], " ~ ", Predictors_1[i], " + ", Predictors_2[i], " + ", Predictors_3[i], " + ", Predictors_4[i]))
  model <- lmer(form, model_data_log_centered)
  
  model_list_osi1 [[i]] <- model
}

tab_model(model_list_osi1  , show.icc = TRUE, digits = 3, show.obs = TRUE, show.re.var = TRUE)

ggplot(model_data_log_centered, aes(adj_Ohdg_TR ,adj_Cd1000_TR)) +  
  geom_point(na.rm = TRUE) +
  geom_smooth(formula = y ~ x,method = "lm", se = FALSE)

ggplot(model_data_log_centered, aes(adj_Ohdg_TR ,adj_Pb1000_TR)) +  
  geom_point(na.rm = TRUE) +
  geom_smooth(formula = y ~ x,method = "lm", se = FALSE)

ggplot(model_data_log_centered, aes(adj_Mda1000_TR ,adj_Cd1000_TR)) +  
  geom_point(na.rm = TRUE) +
  geom_smooth(formula = y ~ x,method = "lm", se = FALSE)

ggplot(model_data_log_centered, aes(adj_Mda1000_TR ,adj_Pb1000_TR)) +  
  geom_point(na.rm = TRUE) +
  geom_smooth(formula = y ~ x,method = "lm", se = FALSE)

ggplot(model_data_log_centered, aes(adj_Isopgf2a1000_TR ,adj_Cd1000_TR)) +  
  geom_point(na.rm = TRUE) +
  geom_smooth(formula = y ~ x,method = "lm", se = FALSE)

ggplot(model_data_log_centered, aes(adj_Isopgf2a1000_TR ,adj_Pb1000_TR)) +  
  geom_point(na.rm = TRUE) +
  geom_smooth(formula = y ~ x,method = "lm", se = FALSE)
```

#### Raw measurements of 8-iso-PGF2a/MDA/8-OHdG and creatinine as fixed effect

```{r set3}
# set3
models <- tibble(Response     = c("Ohdg_TR","Mda1000_TR","Isopgf2a1000_TR","Ohdg_TR","Mda1000_TR","Isopgf2a1000_TR"),
                 Predictors_1 = c("Ohdg_BL + creatinine","Mda1000_BL + creatinine","Isopgf2a1000_BL + creatinine","Ohdg_BL + creatinine",
                                  "Mda1000_BL + creatinine","Isopgf2a1000_BL + creatinine"),
                 Predictors_2 = c("Cd1000_TR","Cd1000_TR","Cd1000_TR","Pb1000_TR","Pb1000_TR","Pb1000_TR"),
                 Predictors_3 = "Sex + age_baseline + DaysfromTreatment",
                 Predictors_4 = "(1 | ID)")

model_list_osi2 <- list()


for (i in 1:nrow(models)) {
  # formula with interaction term
  form <- as.formula(models %$% str_c(Response[i], " ~ ", Predictors_1[i], " + ", Predictors_2[i], " + ", Predictors_3[i], " + ", Predictors_4[i]))
  model <- lmer(form, model_data_log_centered)
  
  model_list_osi2[[i]] <- model
}

tab_model(model_list_osi2 , show.icc = TRUE, digits = 3, show.obs = TRUE, show.re.var = TRUE)

model_list_osi <- append(model_list_osi1, model_list_osi2)

# fdr adjustment taking in account all models for second objective
model_list_osi %>% 
map(~tidy(.)) %>% bind_rows() %>% filter(!term=="(Intercept)") %>%  filter(!str_detect(term, "sd")) %>% 
  mutate(p.adjust = p.adjust(p.value,method="BH"),.after=p.value) %>% 
  mutate(Response = c("adj_Ohdg_TR","adj_Ohdg_TR","adj_Ohdg_TR","adj_Ohdg_TR","adj_Ohdg_TR",
                      "adj_Mda1000_TR", "adj_Mda1000_TR", "adj_Mda1000_TR", "adj_Mda1000_TR","adj_Mda1000_TR",
                      "adj_Isopgf2a1000_TR", "adj_Isopgf2a1000_TR", "adj_Isopgf2a1000_TR", "adj_Isopgf2a1000_TR","adj_Isopgf2a1000_TR",
                      "adj_Ohdg_TR","adj_Ohdg_TR","adj_Ohdg_TR","adj_Ohdg_TR","adj_Ohdg_TR",
                      "adj_Mda1000_TR", "adj_Mda1000_TR", "adj_Mda1000_TR", "adj_Mda1000_TR","adj_Mda1000_TR",
                      "adj_Isopgf2a1000_TR", "adj_Isopgf2a1000_TR", "adj_Isopgf2a1000_TR", "adj_Isopgf2a1000_TR","adj_Isopgf2a1000_TR",
                      "Ohdg_TR","Ohdg_TR","Ohdg_TR","Ohdg_TR","Ohdg_TR","Ohdg_TR",
                      "Mda1000_TR","Mda1000_TR","Mda1000_TR","Mda1000_TR","Mda1000_TR","Mda1000_TR",
                      "Isopgf2a1000_TR","Isopgf2a1000_TR","Isopgf2a1000_TR","Isopgf2a1000_TR","Isopgf2a1000_TR","Isopgf2a1000_TR",
                      "Ohdg_TR","Ohdg_TR","Ohdg_TR","Ohdg_TR","Ohdg_TR","Ohdg_TR",
                      "Mda1000_TR","Mda1000_TR","Mda1000_TR","Mda1000_TR","Mda1000_TR","Mda1000_TR",
                      "Isopgf2a1000_TR","Isopgf2a1000_TR","Isopgf2a1000_TR","Isopgf2a1000_TR","Isopgf2a1000_TR","Isopgf2a1000_TR"
                      ),.before=term) %>%  
  arrange(p.adjust) %>%
  select(!group) %>% 
  kable()

ggplot(model_data_log_centered, aes(Ohdg_TR ,Cd1000_TR)) +  
  geom_point(na.rm = TRUE) +
  geom_smooth(formula = y ~ x,method = "lm", se = FALSE)

ggplot(model_data_log_centered, aes(Ohdg_TR ,Pb1000_TR)) +  
  geom_point(na.rm = TRUE) +
  geom_smooth(formula = y ~ x,method = "lm", se = FALSE)

ggplot(model_data_log_centered, aes(Mda1000_TR ,Cd1000_TR)) +  
  geom_point(na.rm = TRUE) +
  geom_smooth(formula = y ~ x,method = "lm", se = FALSE)

ggplot(model_data_log_centered, aes(Mda1000_TR ,Pb1000_TR)) +  
  geom_point(na.rm = TRUE) +
  geom_smooth(formula = y ~ x,method = "lm", se = FALSE)

ggplot(model_data_log_centered, aes(Isopgf2a1000_TR ,Cd1000_TR)) +  
  geom_point(na.rm = TRUE) +
  geom_smooth(formula = y ~ x,method = "lm", se = FALSE)

ggplot(model_data_log_centered, aes(Isopgf2a1000_TR ,Pb1000_TR)) +  
  geom_point(na.rm = TRUE) +
  geom_smooth(formula = y ~ x,method = "lm", se = FALSE)
```


```{r session, include=FALSE}
# Session information
sessionInfo()
```